nextflu: Real-time tracking of seasonal influenza H3N2 virus evolution in humans

### 2015-03-19 manuscript/nextflu.tex
took a first pass at the ms

%\date{\today}
\maketitle
\bibliographystyle{natbib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
\section{Motivation:}
\section{Results:}
\section{Availability:} \url{http://nextflu.org} and \url{https://github.com/blab/nextflu}.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[bhtp]
	\begin{center}
%\includegraphics[width=0.99\columnwidth]{figures/}
\caption[labelInTOC]{.}
\label{fig:gspace}
\end{center}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%\date{\today}
\maketitle
\bibliographystyle{plain}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract} \section{Motivation:} Seasonal influenza viruses evolve
rapidly and change their antigenic properties. This allows the virus
population to evade immunity in their human hosts and reinfect previously
infected individuals. Similarly, vaccines against seasonal influenza need to
be updated frequently to protect against evolving virus population.

\section{Results:} We have developed a processing pipeline and web-browser
based visualization that allows convenient exploration and analysis of the
most recent influenza virus sequence data. This web-application is live at
\url{nextflu.org} and displays a phylogenetic tree that can be decorated with
several layers of additional information such the viral genotype at specific
sites, sampling location, and sequence or tree based statistics that have been
shown to predictive of future virus dynamics.

\section{Availability:} \url{http://nextflu.org} and \url{https://github.com/blab/nextflu}.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

nextflu consists of a processing pipeline writtin in python and a javascript
based visualization called auspice. As input, augur requires a fasta file of
sequences of the HA segment with fasta labels containing relevant information
such as the strain name, the sampling date and passage history. In a first
step, viruses without complete date  or geographic information and viruses
passaged in eggs are removed. In addition, local outbreaks are filtered by
keeping only one instance of identical sequences sampled at the same location
on the same day.  Following filtering, the viruses are subsampled to achieve
an appoximately even temporal and global distribution (typically 50 viruses
per month). For our standard display period of 3 years, this typically results
in \~1800 viruses, which we align us \texttt{mafft} \citep{mafft}.

Once aligned, the set of virus sequences is further cleaned by removing
insertions relative  to the outgroup and by removing sequences that are either
much closer or further from the outgroup then expected given their date of
sampling (filtering viruses that don't follow an approximate molecular clock).
REASSORTANTS.

From the filtered and cleaned alignment, augur builds a phylogenetic tree using
\texttt{fasttree} \citep{Price:2009p47657} which is further refined by one hour
of RAxML optimization \citep{RAxml}. Next, the state of every internal node
of the tree is inferred using marginal maximum likelihood method and missing 
sequence is filled with the nearest ancestral sequence. Internal branches 
without mutations are collapsed into polytomies. The final tree is decorated
with the attributes to be displayed in the browser, see below.

\begin{figure}[t!]
	\begin{center}
	\includegraphics[width=0.99\columnwidth]{figures/tree_screenshot}
\caption[]{The \texttt{nextflu} website with the user interface on the left and
the tree on the right.}
\label{fig:tree}
\end{center}
\end{figure}

In addition to the phylogenetic tree, augur estimates the frequency
trajectories of mutations, genotypes and clades in the tree. Frequency
trajectories are  represented as a linear interpolation $x(t)$ of values $x_i$
at time points  $t_i$ spaced one month apart. The pivot frequencies
corresponding to a feature $\phi$ (e.g. mutation) are estimated by minimizing
\begin{equation}
\label{eq:freq}
	\begin{split}
	-\log LH(\{x_i\} | \phi)  =& \sum_v I(v,\phi)\log(x(t_v)) + (1-I(v,\phi))\log(1-x(t_v)) \\
			&+\sum_i \frac{(\Delta x_i - \epsilon\Delta x_{i-1})^2}{2\gamma \delta t_i}
\end{split}
\end{equation}
where $I(v,\phi)=1$ if virus $v$ carries attribute $\phi$ and $I(v,\phi)=0$
otherwise. $\Delta x_i = x_i-x_{i-1}$, $\Delta t_i = t_i-t_{i-1}$, $\gamma$
parameterizes the smoothing imposed on the frequency estimate (related to the
frequency changes by genetic drift) and $\epsilon$ parameterizes our
expectation of frequency changes in interval $i$ based on the observed change
in interval $i-1$. $\epsilon=0$ implies that frequency changes are
uncorrelated from one interval to the next, while $\epsilon=1$ implies that
the most likely $\Delta x_i$ is $\Delta x_{i-1}$. We typically use $\epsilon =
0.7$. \EQ{freq} is minimized using SciPy's implementation of a downhill
simplex algorithm \citep{Oliphant:2007p25672}. To speed up convergence,
minization is first done using a coarse grid of pivot points which is
subsequently refined to 12 pivots per year. PUT IN DETAILS ON HOW WE ESTIMATE 
FREQUENCIES ON THE TREE?

The tree decorated with attributes and clade frequencies is written to file in
a nested json format for display in the browser by auspice. In addition to the
tree, json files containing the sequences, frequencies of mutations and
genotypes, as well as meta information are exported.

The tree is visualized by auspice using d3 \citep{d3}. The user can explore
the data interactively by selecting viruses from different dates (moving the
date slider, top right corner in \FIG{tree}) or by coloring the tree by the
following attributes:
\begin{itemize}
	\item epitope mutations: each tip is colored by the number of aminoacid mutations at \~50 
		  positions relative to the root of the tree. These positions are known to
		  be frequent targets of antibodies \citep{shih_simultaneous_2007} and have been
		  suggested to be predictive of future success \citep{luksza_predictive_2014}.
	\item non-epitope mutations: color by the number of mutations outside the above
		  mentioned epitope sites. Mutations outside epitopes tend to be damaging and 
		  anti-correlated with clade expansion \citep{luksza_predictive_2014}.
    \item receptor binding mutations: color by the number of mutations at 7 positions
    	  close to the receptor binding site that have been shown to be responsible 
    	  for major antigenic positions in the past decades \citep{koel_substitutions_2013}.
    \item local branching index is the exponentially weighted tree length surrounding 
    	  a node, which is associated with rapid branching and expansion of clades 
    	  \citep{neher_predicting_2014}. It is recalculated everytime the date slider is
    	  moved and based only on the highlighted viruses.
    \item geographic location
    \item HA1 genotype: entering amino acid positions (separated by a comma) in the 
    	  text field will color every node according to the genotype at these positions.
\end{itemize}

The frequency plot below the tree (see \FIG{freq}) displays the frequency trajectory
of clades in the tree whenever the mouse hovers above the branch defining the clade. 
Furthermore, trajectories of individual mutations, combinations of two mutations, and 
predifined clades such as 3c3.a can be plotted by entering them in text field
\begin{figure}[bhtp]
	\begin{center}
	\includegraphics[width=0.99\columnwidth]{figures/frequencies}
\caption[]{The frequency diagram on \url{nextflu.org} allows geography specific
		plotting of frequencies of individual mutations, pairs of mutations, 
		predefined genotypes, as well as clades in the tree. }
\label{fig:freq}
\end{center}
\end{figure}

\section{Conclusion}
We built nextflu.org to facilitate the analysis and exploration of seasonal influenza
sequence data collected by laboratories around to world. By using the most recent data
and integrating phylogenies with frequency trajectories and sequence or tree based
predictors of successful clades, we hope that nextflu can inform the choice of  
strains used in seasonal influenza vaccines.

nextflu was designed to be readily adapted to other rapidly evolving viruses. 
To that end, the generic classes used for processing the sequence data to include features 
specific to the virus in question.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliography{nextflu}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
